---
title: "Landstingen i siffror"
output: html_notebook
---

##Aktivera programmen som anv?nds

```{r}
library(readxl)
library(tidyverse)
library(skimr)
library(janitor)
```


##Indata

```{r}
library(readxl)
strukturjusterad_kostnad_pp <- read_excel("/Users/ja-p 1/Dropbox/1_dropbox_polsek/landstingsstatistik/vis_export_strukturjusterad_kostnad_per_person.xlsx")

clean_strukturjusterad_kostnad_pp <- strukturjusterad_kostnad_pp %>%
  select("M?tperiod", "L?nskod", "Enhetsnamn", "V?rde") %>%
  rename(landstingskod = "L?nskod", landsting = Enhetsnamn, hs_kostnad_pp = "V?rde")


landstingsskatt <- read_excel("/Users/ja-p 1/Dropbox/1_dropbox_polsek/landstingsstatistik/Landstingsskattesats_SCB_2008_2018.xlsx", n_max = 21, 
    skip = 2)
View(landstingsskatt)

long_landstingsskatt <- landstingsskatt %>%
  gather("2008":"2018", key = "M?tperiod", value = "skatt_%")


```


##S?tt ihop dataset

```{r}

dt <- clean_strukturjusterad_kostnad_pp %>%
  left_join(long_landstingsskatt, by = c("M?tperiod", "landstingskod")) %>%
  select(-landsting.y) %>%
  rename(landsting = "landsting.x")


```

##Hur ser sambandet ut mellan skatt och vad man spenderar? Och hur utvecklas HS-kostnaderna ?ver tid?

```{r}
filter(dt, landsting != "Gotland") %>%
  ggplot(aes(x = `skatt_%`, y = hs_kostnad_pp)) + 
  geom_point()

filter(dt, landsting != "Gotland") %>%
  ggplot(aes(x = M?tperiod, y = hs_kostnad_pp, group = landsting)) +
  geom_line(aes(color=`skatt_%`))
  
 

```




Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
